---
title: "scRNA-seq data preprocessing"
output: github_document
---

## loading libraries
```{r libraries}
library(Seurat)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
```

## read in scRNA-seq data
```{r seurat object}
#load RNA data with ReadMtx function from Seurat package
counts_matrix <- ReadMtx(
  "/home/fgsasse_lrs_1/Downloads/BA/BA_data/RNA/matrix.mtx",
  "/home/fgsasse_lrs_1/Downloads/BA/BA_data/RNA/cells.tsv",
  "/home/fgsasse_lrs_1/Downloads/BA/BA_data/RNA/features.tsv",
  cell.column = 1,
  feature.column = 1,
  cell.sep = "\t",
  feature.sep = "\t",
  skip.cell = 0,
  skip.feature = 0,
  mtx.transpose = TRUE,
  unique.features = TRUE,
  strip.suffix = FALSE
)

seu.obj <- CreateSeuratObject(counts = counts_matrix)
#saveRDS(seu.obj, file = "seurat_obj.rds")
```

# adding metadata to seurat object
```{r seurat object metadata}
#adding metadata to seurat object
metadata_old <- seu.obj@meta.data
metadata_old$cell_id <- rownames(metadata_old)

metadata_new <- read.delim("/home/fgsasse_lrs_1/Downloads/BA/BA_data/RNA/cells.tsv", header = FALSE)

colnames <- c("cell_id",'developmental_stage', 'dataset', 'zebrafish_anatomy_ontology_class',
       'zebrafish_anatomy_ontology_class_coarse', 'timepoint',
       'n_genes_by_counts', 'total_counts', 'total_counts_mt', 'pct_counts_mt',
       'total_counts_nc', 'pct_counts_nc')
colnames(metadata_new) <- colnames
head(metadata_new)

metadata <- merge.data.frame(metadata_old, metadata_new, by = "cell_id", sort=FALSE)
rownames(metadata) <- metadata$cell_id

all(rownames(metadata) == rownames(seu.obj@meta.data)) # check if rownames of metadata match colnames of seurat object
seu.obj@meta.data <- metadata
```


## QC and filtering
# comparing the number of cells in each cell type (zebrafish anatomy ontology class) with and without the developmental stages

```{r QC filtering}
#generating df of original data set
so_df <- seu.obj@meta.data

unique(so_df$timepoint)
unique(so_df$zebrafish_anatomy_ontology_class )
unique(so_df$dataset) #fish samples

#numbers of cells in each cell type (= zebrafish anatomy ontology class)
cells_celltype1 <- so_df %>% group_by(zebrafish_anatomy_ontology_class) %>% summarise(n_cells = n())
cells_celltype1

ct_n <- ggplot(cells_celltype1, aes(x=zebrafish_anatomy_ontology_class, y= n_cells, fill = zebrafish_anatomy_ontology_class))+
   geom_col(alpha = 0.7) +
   geom_text(aes(label = ifelse(n_cells == max(n_cells) | n_cells == min(n_cells), n_cells, "")),
             vjust = -0.4, size = 2.5, col = "darkblue", fontface = "bold") +
   theme_bw()+ 
   theme(axis.text.x = element_text(angle = 70, hjust = 1), legend.position = "none") +
   labs(title = "Number of cells in each cell type before filtering",
                     x = "cell type",
                     y = "number of cells")
ct_n

#numbers of cells of every density()#numbers of cells of every developmental stages in each cell type (n_cells#numbers of cells of every developmental stages in each cell type (= zebrafish anatomy ontology class)
cells_time1 <- so_df %>% group_by(zebrafish_anatomy_ontology_class, developmental_stage) %>% summarise(n_cells = n()) 
cells_time1 

ct_t_n <- ggplot(cells_time1, aes(x=zebrafish_anatomy_ontology_class, y= n_cells, fill = developmental_stage))+
   geom_col(alpha = 0.7) +    
  geom_text(aes(label = ifelse(n_cells < 100, n_cells, "")),
             vjust = -0.4, size = 2.5, col = "darkblue", fontface = "bold") +
   facet_wrap(~ developmental_stage, ncol = 1)+ theme_bw(base_size = 6)+
   theme(axis.text.x = element_text(angle = 70, hjust = 1), axis.text = element_text(size = 8)) +
   labs(title = "Number of cells in each developmental stages in each cell type before filtering",
                     x = "cell type",
                     y = "number of cells")
ct_t_n

#number of cells in each developmental stages of each cell type
c <- cells_time1 %>% group_by(zebrafish_anatomy_ontology_class, developmental_stage) %>% mutate(n_below100c = sum(n_cells < 100))
c %>% group_by(n_below100c) %>% count()
```
Regarding the number of cells in each cell type: 
- only 87 cells of "primordial germ" cells (min)
- 5110 of "tail bud" cells (max)

Regarding the number of cells in each developmental stages of each cell type:
- in "15 somites stage" only 2 cell types below 100 cells (min)
- in "30 somites stage" 17 cell types below 100 cells (max)


#Filtering the data by removing every cell type in each developmental stages below 100 cells
```{r plots of un-/filtered data}
#filtering out the celltype & timepoints below 100 cells
filtered <- so_df %>% group_by(zebrafish_anatomy_ontology_class, timepoint) %>% summarise(n_cells = n())%>% mutate(rm = n_cells <100)
filtered <- filtered %>% mutate(keep_cell = ifelse(!rm, paste(zebrafish_anatomy_ontology_class, timepoint), ""))

fil_cells <- filtered$keep_cell #cells we are keeping 

so_df$cell_name <- paste(so_df$zebrafish_anatomy_ontology_class, so_df$timepoint)

filtered_df <- so_df %>% filter(cell_name %in% fil_cells)

#comparing the filtered data with unfiltered data by plotting
cells_celltype2 <- filtered_df %>% group_by(zebrafish_anatomy_ontology_class) %>% summarise(n_cells = n())
cells_celltype2

ct_n_fil <- ggplot(cells_celltype2, aes(x=zebrafish_anatomy_ontology_class, y= n_cells, fill = zebrafish_anatomy_ontology_class))+
   geom_col(alpha = 0.7) +
   geom_text(aes(label = ifelse(n_cells == max(n_cells) | n_cells == min(n_cells), n_cells, "")),
             vjust = -0.4, size = 2.5, col = "darkblue", fontface = "bold") +
   theme_bw()+ 
   theme(axis.text.x = element_text(angle = 70, hjust = 1), legend.position = "none") +
   labs(title = "Number of cells in each cell type after filtering",
                     x = "cell type",
                     y = "number of cells")
ct_n| ct_n_fil 

#numbers of cells of every density()#numbers of cells of every developmental stages in each cell type (n_cells#numbers of cells of every developmental stages in each cell type (= zebrafish anatomy ontology class)
cells_time2 <- filtered_df %>% group_by(zebrafish_anatomy_ontology_class, developmental_stage) %>% summarise(n_cells = n()) 
cells_time2 

ct_t_fil_n <- ggplot(cells_time2, aes(x=zebrafish_anatomy_ontology_class, y= n_cells, fill = developmental_stage))+
   geom_col(alpha = 0.7) +    
  geom_text(aes(label = ifelse(n_cells < 100, n_cells, "")),
             vjust = -0.4, size = 2.5, col = "darkblue", fontface = "bold") +
   facet_wrap(~ developmental_stage, ncol = 1)+ theme_bw(base_size = 6)+
   theme(axis.text.x = element_text(angle = 70, hjust = 1), axis.text = element_text(size = 8)) +
   labs(title = "Number of cells in each developmental stages in each cell type after filtering",
                     x = "cell type",
                     y = "number of cells")
ct_t_n | ct_t_fil_n 

#creating new Seurat object by filtered cell_ids
cell_id_fil <- filtered_df$cell_id

fil.seu.obj <- subset(seu.obj, cell_id %in% cell_id_fil)

```
#plots of original and filtered data set regarding mt percentage and total counts
plot_mt <- ggplot(so_df, aes(x = pct_counts_mt, after_stat(scaled)))+
            geom_density() +
            geom_density(aes(color = fil_cell), alpha = 0.3)+
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(title = "Distribution of mitochondrial gene expression (%)  ",
                     x = "Percentage of mitochondrial gene expression",
                    y = "Number of cells (%)")
plot_mt


plot_tc <- ggplot(so_df, aes(x = total_counts, after_stat(scaled)))+
            geom_density() + 
            geom_density(aes(color = fil_cell)) +
            scale_x_log10() +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(title = "Distribution of total counts",
                     x = "Total counts",
                    y = "Number of cells (%)")
plot_tc

plot_tc_mt <- ggplot(so_df, aes(x = total_counts, y = pct_counts_mt))+
            geom_point(aes(col = fil_cell), alpha = 0.01, size = 0.5)+
            geom_smooth(method = "lm", col = "black", linewidth = 0.5,  se = FALSE)+
            scale_x_log10() + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(title = "Total counts vs percentage of mitochondrial gene expression ",
                     x = "Total counts",
                    y = "Percentage of mitochondrial gene expression")
plot_tc_mt 




The number of cells in each cell type decreases insanely after filtering out as shown in the plots.
Next step: See if marker genes of one cell type can be still detected after filtering
- barcode ids of cells before and after filtering
- gene marker analysis


## Seurat's standard workflow with filtered data
#Before continuing with the following analysis, we'll visualize the original data first
```{r Plots regarding developmental stages and cell types}
mycol <- c("#d9ed92", "#b5e48c", "#99d98c", "#76c893", "#52b69a", "#34a0a4", "#168aad", "#1a759f", "#1e6091", "#184e77" )
mycol2 <- c("#54478C", "#2C699A", "#048BA8", "#0DB39E", "#16DB93", "#83E377", "#B9E769", "#EFEA5A", "#F1C453", "#F29E4C")


#Visualizing the QC metrics regarding.. 
#whole dataset
VlnPlot(
  fil.seu.obj,
  features = c("nFeature_RNA", "nCount_RNA", "pct_counts_mt"),
  ncol = 3,
  alpha = 0.1,
  pt.size = 0,
  cols = mycol
) 

#developmental stages
VlnPlot(
  fil.seu.obj,
  features = c("nFeature_RNA", "nCount_RNA", "pct_counts_mt"),
  ncol = 3,
  group.by = "timepoint",
  alpha = 0.1,
  pt.size = 0,
  cols = mycol
) + theme(legend.position = "right") + 
  scale_fill_discrete(name= "developmental stages") &
  theme( axis.text.x = element_blank()) 

#cell types
VlnPlot(
  fil.seu.obj,
  features = c("nFeature_RNA", "nCount_RNA", "pct_counts_mt"),
  ncol = 3,
  group.by = "zebrafish_anatomy_ontology_class",
  alpha = 0.1,
  pt.size = 0
) +  
  theme(legend.position = "right") + 
  scale_fill_discrete(name= "cell type") &
  theme( axis.text.x = element_blank()) 

#Scatterplot show good regression between genes (unique genes detected) and counts (total UMI counts) across timepoints
FeatureScatter(fil.seu.obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size = 0.00001, group.by = "developmental_stage") + 
  geom_point(aes(color = "developmental_stage"), size = 0.1, alpha = 0.01) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() 

  #the higher nFeature the nCounts should be, cause high nCounts represents higher sequencing depth

#Scatterplot 
FeatureScatter(fil.seu.obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size = 0.00001, group.by = "zebrafish_anatomy_ontology_class") + 
  geom_point(aes(color = "zebrafish_anatomy_ontology_class"), size = 0.1, alpha = 0.01) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() 

#ggplot(so_df, aes(x = nCount_RNA, y = nFeature_RNA)) +
  #geom_point(aes( color = developmental_stage), alpha = 0.01, size = 1) +
  #geom_smooth(method = "lm")+
  #theme(axis.text.x = element_text(angle = 70, hjust = 1))+
 # scale_color_manual(values = mycol) + theme_bw()

#ggplot(so_df, aes(x = nCount_RNA, y = nFeature_RNA, color = developmental_stage)) +
  #geom_point(alpha = 0.01, size = 0.5) +
  #facet_wrap(~ developmental_stage, nrow = 2, shrink = FALSE) +
  #geom_smooth(method = "lm", col = "black", linewidth = 0.5,  se = FALSE)+ theme_bw() +
  #theme(axis.text.x = element_text(angle = 70, hjust = 1))+
```


# Seurat workflow
1. Normalization
The standard normalization method in Seurat is a global-scaling normalization method (NormalizeData(), default LogNormalize), which transforms the feature counts for each cell by dividing by the total counts of the cell and multiplying by a scale factor (default = 10000) and then applying a natural log with a pseudocount of 1. This should help reduce variation from sequencing depth and somewhat prevent high abundance genes from dominating downstream analyses.
- we scaled the scale.factor = 1e6

2. Finding Variable Features (marker genes)
A subset of cells with high cell to cell variation can be selected using the variance stabilization transformation method of FindVariableFeatures(). This uses a mean-variance relationship to return 2,000 variable features to prioritize for downstream analyses. 

3. Scaling the data
As a standard preprocessing step prior PCA, a linear transformation ('scaling') is applied with ScaleData().
The ScaleData() function:
- Shifts the expression of each gene, so that the mean expression across cells is 0
- Scales the expression of each gene, so that the variance across cells is 1
=> This step gives equal weight in downstream analyses, so that highly-expressed genes do not dominate

4. Performing linear Dimension Reduction via PCA - determining resolution & dimension
- maximizes variance (Global structure)

5. Clustering based on the PCA rotations

6. Performing non-linear dimension reduction via UMAP
- preserves local relationship

```{r Seurat workflow}
#Normalization of data: ((gene expression of each cell/total expression)*scale factor (10000)) -> log-transformation; log()
fil.seu.obj <- NormalizeData(fil.seu.obj, normalization.method = "LogNormalize", scale.factor = 1e6)


#Identification of high variable features (= genes) with "FindVariableFeatures()", which shows a high cell-to-cell variation
fil.seu.obj <- FindVariableFeatures(fil.seu.obj, selection.method = "vst")
top10genes <- head(VariableFeatures(fil.seu.obj),10)

plot_vf <- VariableFeaturePlot(fil.seu.obj)
LabelPoints(plot = plot_vf, points = top10genes, repel = TRUE)

#Identification of clusters (= cell population) with its differential expression between the clusters by "FindAllMarkers()"

#Scaling of the data to mitigate "irrelevant" sources of variation
#all_genes <- rownames(seu.obj.fil)
fil.seu.obj <- ScaleData(fil.seu.obj)

#Linear dimension reduction: PCA & determining dimensionality of the data
fil.seu.obj <- RunPCA(fil.seu.obj, features = VariableFeatures(fil.seu.obj))
ElbowPlot(fil.seu.obj)

#Clustering 
fil.seu.obj <- FindNeighbors(fil.seu.obj, dims = 1:18)
fil.seu.obj <- FindClusters(fil.seu.obj, resolution = c(0.25, 0.5, 0.75, 1))
DimPlot(fil.seu.obj, group.by = "RNA_snn_res.0.5", label = TRUE)

#Non-linear dimension reduction: UMAP 
fil.seu.obj <- RunUMAP(fil.seu.obj, dims = 1:18)


#saveRDS(fil.seu.obj, file = "/home/fgsasse_lrs_1/Downloads/BA/so_preprocessed")
#markers <- FindAllMarkers(seu.obj.fil) #ups dauert laenger als gedacht..
ct_plot <- DimPlot(fil.seu.obj, reduction = "umap", group.by = "zebrafish_anatomy_ontology_class", label = TRUE)
timepoint_plot <- DimPlot(fil.seu.obj, reduction = "umap", group.by = "timepoint", label = TRUE)

ct_plot|timepoint_plot

```



## Pseudobulking scRNA-Seq data
After looking in to the single cell RNA data, we create a pseudobulk RNA count matrix with raw counts grouped by cell type & developmental stages.
DE analysis can be performed by the function FindMarkers() but these tests treat each cell as an independent replicate and ignore inherent correlations between cells originating from the same sample. Such analyses have been shown to find a large number of false positive associations, as has been demonstrated by Squair et al., 2021, Zimmerman et al., 2021, Junttila et al., 2022, and others. 
Pseudobulking can be used to account for such within-sample correlation. To pseudobulk, we will use AggregateExpression() to sum together gene counts of all the cells from the same sample for each cell type. This results in one gene expression profile per sample and cell type. We can then perform DE analysis using DESeq2 on the sample level. This treats the samples, rather than the individual cells, as independent observations. 
-> for average expression of gene counts, AverageExpression()
-> one embryo for each time point but two embryos for 16hpf (embryo = dataset)

```{r Pseudobulking}
#Pseudobulking the counts from the same sample (fish) for each cell type 
#fil.seu.obj <- readRDS(file = "/home/fgsasse_lrs_1/Downloads/BA/BA_data/Rmd/seurat_obj.rds")
#pseudo_ct_seu.obj.fil <- AggregateExpression(fil.seu.obj,
                                             #group.by = c("zebrafish_anatomy_ontology_class"),
                                             #assays = "RNA",
                                             #slot = "data",
                                            # return.seurat = TRUE)

#Pseudobulking the counts from the same sample for each cell type across all developmental stages
#pseudo_all_seu.obj.fil <- AggregateExpression(fil.seu.obj, 
                              #group.by = c("zebrafish_anatomy_ontology_class", "timepoint"),
                              #assays = "RNA",
                              #slot = "data",
                              #return.seurat = FALSE)

#Idents(pseudo_all_seu.obj.fil) <- "zebrafish_anatomy_ontology_class"

#head(FindMarkers(pseudo_all_seu.obj.fil, ident.1 = "differentiating-neurons", ident.2 = NULL ),20)

#saving the pseudobulk matrices as MatrixMarket-format 
#bulk_ct_counts <- pseudo_ct_seu.obj.fil$RNA

#psudobulk_dir <- "/home/fgsasse_lrs_1/Downloads/BA/BA_data/Pseudobulks/RNA/celltypes_times"
                  #Matrix::writeMM(bulk_ct_counts, file.path(psudobulk_dir, "matrix.mtx"))
                  #write.table(colnames(bulk_ct_counts),
                             # file.path(psudobulk_dir, "barcodes.tsv"),
                              #quote = FALSE, row.names = FALSE, col.names = FALSE)
                 # write.table(rownames(bulk_ct_counts),
                              #file.path(psudobulk_dir, "features.tsv"),
                              #quote = FALSE, row.names = FALSE, col.names = FALSE)

#bulk_all_counts <- pseudo_all_seu.obj.fil$RNA
#psudobulk_dir2 <- "/home/fgsasse_lrs_1/Downloads/BA/BA_data/Pseudobulks/RNA/celltypes_times"
                  #Matrix::writeMM(bulk_ct_counts, file.path(psudobulk_dir2, "matrix.mtx"))
                  #write.table(colnames(bulk_ct_counts),
                             # file.path(psudobulk_dir2, "barcodes.tsv"),
                              #quote = FALSE, row.names = FALSE, col.names = FALSE)
                  #write.table(rownames(bulk_ct_counts),
                              #file.path(psudobulk_dir2, "features.tsv"),
                             # quote = FALSE, row.names = FALSE, col.names = FALSE)

```
New idea for pseudobulking:
Geometric mean is more robust to outliers than arithmetic mean.
- Arithmetic mean is the sum of the values divided by the total number of values = the average of the values
- Geometric mean is the product of the values raised to the multiplicative inverse of the total number of values

#Pseudobulking RNA 
```{r}
#Retraining the raw counts as matrix 
mat <- fil.seu.obj@assays$RNA$counts
mat <- t(mat)
head(mat)

#Creating a df of meta data
meta = as.data.frame(fil.seu.obj@meta.data)
head(meta)

#aligning the rownames of the matrix as same as the meta data (they should be in the same order) and creating a function how to aggregate the RNA counts: geometrical mean/ arithmetric mean

rownames(mat) <- meta$cell_id

split(meta$cell_id, meta$zebrafish_anatomy_ontology_class) #now only separated in cell types

agg_fun <- function(mat, meta, fun="mean") {
  if (fun == "mean") {
    agg_f <- function(x) {
      colSums(mat[x, ]) / sum(mat[x,])
    }
  } else if (fun == "geom_mean") {
    agg_f <- function(x) {
      exp(colMeans(log(mat[x,])))
    }
  }
  t(sapply(split(meta$cell_id, meta$zebrafish_anatomy_ontology_class), agg_f))
}

#aggregating by geometrical mean
geom_agg_rna <- agg_fun(mat, meta, fun = "geom_mean")
head(geom_agg_rna)
max(geom_agg_rna)

#aggregating by arithmetrical mean
arth_agg_rna <- agg_fun(mat, meta, fun = "mean")
head(arth_agg_rna)
max(arth_agg_rna)

#dim(39 32057)

  #abs(agg_mean - agg_geom)

```



```{r}
# create combined grouping
group <- paste(meta$zebrafish_anatomy_ontology_class,
                 meta$developmental_stage,
                 sep = "_")
  
#split CELL IDS by group
split_cells <- split(meta$cell_id, group)

rownames(mat) <- meta$cell_id

agg_fun2 <- function(mat, meta, fun="mean") {
  if (fun == "mean") {
    agg_f <- function(x) {
      colMeans(mat[x, ])
    }
  } else if (fun == "geom_mean") {
    agg_f <- function(x) {
      exp(colMeans(log(mat[x,])))
    }
  }
  t(sapply(split_cells, agg_f))
}

rownames(mat) <- meta$cell_id

geom_aggall_rna <- agg_fun2(mat, meta, fun = "geom_mean")

head(geom_aggall_rna)
max(geom_aggall_rna)
#dim(196 32057)

geom_aggall_rna <- agg_fun2(mat, meta, fun = "geom_mean")

head(mean_aggall_rna)
max(mean_aggall_rna)
#dim(196 32057)
```


```{r}
writeLines(cell_id_fil, "cell_id.txt")
```


